Numerical modelling of the Cordilleran ice sheet

### 2014-07-20 kappa.tex
Add more corrections from Arjen.

As a first step towards a simulations of the Cordilleran ice sheet through the
last glacial cycle, I have opted to run the ice sheet model under a range of
constant temperature depressions, targeting the LGM as the period for which ice
sheet limits are best known. These simulations, presented in \CCLI\ and
performed on a 10\,km-resolution grid, use constant temperature depressions
from the present climate state, which is represented using five different input
climatologies. All model runs start
from ice free conditions and run for 10\,ka, a period chosen as representative
of Cordilleran ice sheet build-up from assumed, negligible ice cover. Although
this assumption appears unjustified from results in \CCYC, the simplistic
set-up used in \CCLI\ allowed an assessment of model sensitivity to the choice
of input climatologies, without involving too many additional parameters.

Discrepancies between the five input data sets used (\cref{tab:datasources})
caused considerable variability in modelled ice sheet geometries (\CCLI,
Figs.~5--7). In particular, ice growth in continental sectors of the modelling
domain is strongly dependent on input precipitation patterns, which are often
poorly resolved by atmospheric reanalysis fields used to force the model
(\CCLI, Figs.~10--14). The closest match between modelled ice sheet extent and
the reconstructed LGM ice margin \citep{Dyke.2004} was obtained using the North
American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006} climatologies
and a temperature depression of 7\,K (\CCLI, Fig.~15). The relevant simulation
is reproduced here, using a higher horizontal resolution of 3\,km
(\cref{fig:plot-lgm}). Following this sensitivity study, I have retained the
NARR climatology as input to simulations of the Cordilleran ice sheet presented
here and in \CCYC.

@@

The effect of the PDD standard deviation, $\sPDD$, on simulations of the
Cordilleran ice sheet is not shown in any of the papers. However, following
these two studies, a natural step forward was to implement, in PISM, spatially
and seasonally variable $\sPDD$ input (PISM issue~\issue{179}), and the
parameterisation expressed by \cref{eqn:sigma} over ice-filled grid cells
(PISM issue~\issue{265}). Their effect on modelled Cordilleran ice sheet
surface elevations is demonstrated here using a constant climate set-up similar
to the one used in \CCLI\ (\cref{fig:plot-sdeffect}). These simulations show
that the effect of daily variability is strongest in continental regions, where
the $\sPDD$ values are highest (cf. \cref{fig:plot-sdmap}). Where comparing the
results to the mapped LGM ice limits (\cref{fig:map-cordillera}), which show
a restricted ice extent in the continental areas despite low temperatures at
present (\cref{fig:plot-atm}), it appears that including daily temperature
variability in the ice sheet model results in closer agreement with these
limits (\cref{fig:plot-sdeffect}b and~c). However, applying the parametrisation
derived for the Greenland ice sheet (\cref{eqn:sigma}) over ice-filled grid
cells partly reverts this effect (\cref{fig:plot-sdeffect}d), indicating that
this parametrization may not be applicable to the former Cordilleran ice sheet.

@@

The primary goal of this thesis has been to perform simulations of the
Cordilleran ice sheet through the last glacial cycle. This is the topic of
\CCYC, where monthly air temperature, precipitation, and PDD standard deviation
climatologies from the NARR are combined with palaeo-temperature time-series
from six proxy records to produce a time-dependent climate forcing (\CCYC,
Figs.~2--3). The resulting simulations, using horizontal resolutions of 6 and
10\,km, all produced two major glaciations during the last glacial cycle, at
MIS~4 and MIS~2. However, the differences in timing and magnitude of these
glaciations are considerable (\CCYC, Figs.~3--4).

Geological evidence for the timing and extent of maximum glaciation during the
LGM allowed me to select two simulations, driven by the GRIP
\citep{Dansgaard.etal.1993} and EPICA \citep{Jouzel.etal.2007} ice core records
(\CCYC, Figs.~5--6), as producing the most realistic reconstructions of the
last glacial cycle. Using aggregated model output from these two runs, I
analysed the location of transient ice caps before and after the LGM, the
longevity of major ice-dispersal centres, and the rates and patterns of
deglaciation (\CCYC, Figs.~6--15), in relation to available geological
evidence. Modelled ice volume evolution, expressed as metres of sea level
equivalent, and snapshots of ice extent and flow patterns from the GRIP
simulation (\cref{fig:plot-snapshots}) show the growth and decay of the
Cordilleran ice sheet over its last expansion cycle. Ice sheet growth begins on
a residual ice cap located over the Skeena Mountains, which also corresponds
to the location of the last remnants of the ice sheet towards the end of the
deglaciation.

@@

Through sensitivity studies, I have identified the predominant sources of
uncertainty involved in numerical modelling of Cordilleran ice sheet. Through
the different papers constituting this thesis, I present an assessment of some
of these sources, including the choice of present-day atmospheric forcing, the
effects of daily temperature variability, and the choice of palaeo-temperature
time series used to drive the model through past climate history.

The resulting simulations, combined with available geological evidence,
yielded estimates of the LGM Cordilleran ice sheet extent and volume,
identification of the main ice-dispersal centres througout the last glacial
cycle, and understanding of final deglaciation patterns. However, there remains
considerable room for model improvements and further analysis of the simulation
output. Sources of error not assessed in this thesis include potential
buttressing effects of the Laurentide ice sheet, past changes in precipitation,
and regional variability of the geothermal heat flux.

@@

Certainly the most important assumption made in this thesis is that of a
Cordilleran ice sheet largely disconnected from its much larger neighbour, the
\emph{Laurentide ice sheet}. Geomorphological evidence from the suture zone
indicates that, during the LGM, both Cordilleran and Laurentide ice flow has
been deflected by their junction \citep{Kleman.etal.2010, Margold.etal.2013,
Margold.etal.2013a}, although it remains unclear for how long this junction has
lasted \citep{Jackson.etal.1997, Bednarski.Smith.2007, Gowan.2013}.

Nevertheless, during most of the last glacial cycle, the two ice sheets have
been disconnected \citep{Kleman.etal.2010}. Thus, this assumption only affects
numerical reconstructions of the Cordilleran ice sheet presented here during
the last glacial maximum and during the subsequent deglacation. Other numerical
modelling results \citep{Gregoire.etal.2012} indicate the Laurentide ice sheet
may have formed a buttress to ice flow from the Western Cordillera, resulting
in a thicker Cordilleran ice sheet. However modelling alone will not elucidate
the nature and duration of the interaction between of the Cordilleran and
Laurentide ice sheets. For this, additional field constraints will be needed.

@@

Another important source of error is the treatment of \emph{precipitation}, to
which
the numerical ice sheet model is highly sensitive (\CCLI). In all sensitivity
studies, we base precipitation patterns on current values derived atmospheric
reanalyses. Although the distribution of precipitation has undoubtedly
been affected by past changes in global climate and the intermittent presence
of ice sheets and ice caps in the Northern Hemisphere, its changes are
difficult to assess. In the North American Cordillera, much of the continental
interior was likely dryer than today due to the presence of ice caps on the
Pacific Coast ranges, reinforcing the current topographic barrier. However,
some regions were probably rainier as a side effect of modified wind patterns.

In meteorology, these effects are typically analysed using atmospheric
circulation models. However, I have shown that different atmospheric reanalyses
--- circulation models using data assimilation techniques --- fail to agree
regarding the present-day pattern of precipitation over the modelling domain.
These discrepancies, which arise as a consequence of the highly mountainous
terrain, have dramatic effects on the ice sheet model response. Considering the
lack of observational data for modelling past climates, the increased
computational demand, and a decreased familiarity to processes at play, we must
expect that, for ice sheet simulations forced by different palaeo-climate
simulations, the uncertainty associated to the choice of atmospheric forcing
will be even higher. Perhaps alternative methods of coupling between ice sheet
evolution and precipitation changes need to be developed.

@@

I expect that inclusion of geothermal heat patterns will allow modelled basal
flow directions and distances to be compared with glacial lineations, which is
one of the main sources of geomorphological evidence used to reconstruct former
ice sheets \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}. For
instance, in our simulations, large portions of the Cordilleran ice sheet
remain cold-based throughout the deglaciation (\CCYC, Fig.~15). As cold-based
glaciers typically do not impact their beds significantly, this could indicate
that some of the lineation patterns we observe within these zones of consistent
cold-based deglaciation could be older features, formed during an earlier phase
of advance or retreat of the ice sheet. However, without a better control on
the geothermal heat flux, such analysis are perhaps too premature.

@@

In this thesis, I present numerical reconstructions of the former Cordilleran
ice sheet to a level of detail that has, to my knowledge, not been attained
before. Although mathematically uncertain, the coupled SIA-SSA approximation,
combined with massively parallel computing, allowed ice sheet modelling at
high spatial resolution and over multi-millenial time scales characteristic
of ice sheet growth and decay.

Through sensitivity studies, and comparison of model output to geological
evidence, I have gained a personal understanding of Cordilleran ice sheet
evolution through the last glacial cycle. I attempt to summarise my thoughts
here in a few points, perhaps also applicable to other ice sheets in the
Northern Hemisphere.

@@

\item Although spatial patterns of glaciation in the North American Cordillera
are sometimes paradoxical, they can be largely explained by the present-day
distribution of temperature and precipitation in the region, provided an
accurate treatment of seasonality, orographic precipitation, and daily
variability in the model, and a simple temperature-index accumulation and melt
model.

\item The LGM Cordilleran ice sheet corresponds to a transient stage, out of
balance with the contemporaneous atmospheric temperatures. Its maximum extent
is the result of a 50\,ka-long evolution, captured in the geologic record at a
moment of transition from colder to warmer temperatures than those needed to
sustain a constant ice sheet geometry. Therefore, this maximum stage is not
representative of average conditions of the last glacial cycle.

\item The patterns of growth and decay of the Cordilleran ice sheet are varied
and differ by region and by epoch. Two-dimensional, conceptual models of growth
and decay \citep[e.g.][]{Fulton.1991, Margold.etal.2013a} do not encompass this
diversity. In order to decipher the complexity of geological evidence, the
Cordilleran ice sheet needs to be thought of as a three-dimensional body of
moving ice in perpetual transformation towards evolving climate conditions.

>>>>

As a first step towards a simulations of the Cordilleran ice sheet through the
last glacial cycle, I have opted to run the ice sheet model under a range of
constant temperature depressions, targeting the LGM as the period for which ice
sheet limits are best known. These simulations,
performed on a 10\,km grid, use constant temperature depressions
from the present climate state, which is represented using five different input
climatologies. All model runs start
from ice free conditions and run for 10\,ka, a period chosen as representative
of Cordilleran ice sheet build-up from assumed, negligible ice cover. Although
this assumption appears unjustified from results in \CCYC, the simplistic
set-up used in \CCLI\ allowed an assessment of model sensitivity to the choice
of input climatologies, without involving too many additional parameters.

Discrepancies between the five input data sets used (\cref{tab:datasources})
caused considerable variability in modelled ice sheet geometries (\CCLI,
Figs.~5--7). In particular, ice growth in continental sectors of the modelling
domain is strongly dependent on input precipitation patterns, which are often
poorly resolved by atmospheric reanalysis fields used to force the model
(\CCLI, Figs.~10--14). The closest match between modelled ice sheet extent and
the reconstructed LGM ice margin \citep{Dyke.2004} was obtained using the North
American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006} climatologies
and a temperature depression of 7\,K (\CCLI, Fig.~15). The relevant simulation
is reproduced here, using a higher horizontal resolution of 3\,km
(\cref{fig:plot-lgm}). Following this sensitivity study, I have retained the
NARR climatology as input to simulations of the Cordilleran ice sheet through
the last glacial cycle (\CCYC).

@@

The effect of the PDD standard deviation, $\sPDD$, on simulations of the
Cordilleran ice sheet is not shown in any of the papers. However, following
these two studies, a natural step forward was to implement, in PISM, spatially
and seasonally variable $\sPDD$ input (PISM issue~\issue{179}), and the
parameterisation expressed by \cref{eqn:sigma} over ice-filled grid cells
(PISM issue~\issue{265}). Their effect on modelled Cordilleran ice sheet
surface elevations is demonstrated here using a constant climate set-up similar
to the one used in \CCLI\ (\cref{fig:plot-sdeffect}). These simulations show
that the effect of daily variability is strongest in continental regions, where
the $\sPDD$ values are highest (cf. \cref{fig:plot-sdmap}). When the results
are compared to the mapped LGM ice limits (\cref{fig:map-cordillera}), which show
a restricted ice extent in continental areas despite low temperatures at
present (\cref{fig:plot-atm}), it appears that including daily temperature
variability in the ice sheet model results in a closer agreement with these
limits (\cref{fig:plot-sdeffect}b and~c). However, applying the parametrisation
derived for the Greenland ice sheet (\cref{eqn:sigma}) over ice-filled grid
cells partly reverts this effect (\cref{fig:plot-sdeffect}d), indicating that
this parametrization may not be applicable to the former Cordilleran ice sheet.

@@

The primary goal of my thesis has been to perform numerical reconstructions
of Cordilleran ice sheet history through the last glacial cycle, which forms
the topic of \CCYC.
Here, monthly air temperature, precipitation, and PDD standard deviation
climatologies from the NARR are combined with palaeo-temperature time-series
from six proxy records to produce a time-dependent climate forcing
spanning over the last glacial cycle (\CCYC,
Figs.~2--3). The resulting simulations, using horizontal resolutions of 6 and
10\,km, all produced two major glaciations during the last glacial cycle, at
MIS~4 and MIS~2. However, the differences in timing and magnitude of these
glaciations are considerable (\CCYC, Figs.~3--4).

Geological evidence for the timing and extent of maximum glaciation during the
LGM allowed me to select two simulations, driven by the GRIP
\citep{Dansgaard.etal.1993} and EPICA \citep{Jouzel.etal.2007} ice core records
(\CCYC, Figs.~5--6), as producing the most realistic reconstructions of the
last glacial cycle. Using aggregated model output from these two runs, I
analysed the location of transient ice caps before and after the LGM, the
longevity of major ice-dispersal centres, and the rates and patterns of
deglaciation (\CCYC, Figs.~6--15), in relation to available geological
evidence. Modelled ice volume evolution, expressed as metres of sea level
equivalent, and snapshots of ice extent and flow patterns from the GRIP
simulation (\cref{fig:plot-snapshots}) show the growth and decay of the
Cordilleran ice sheet over its last expansion cycle. An ice cap located over
the Skeena Mountains acts as the primary nucleation center for the Cordilleran
ice sheet. After the LGM the ice sheet shrinks rapidly such that the last
remnants of the ice sheet towards the end of the deglaciation in the early
Holocene reside once again in the Skeena Mountains.

@@

Through sensitivity studies presented in my thesis, I have identified some of
the predominant sources of uncertainty involved in numerical modelling of the
Cordilleran ice sheet, including the choice of present-day atmospheric forcing
fields, the effects of daily temperature variability, and the choice of
palaeoclimate time series used simulate ice sheet growth over glacial cycle
time scales. The resulting simulations, combined with available geological
evidence, yielded estimates of the LGM Cordilleran ice sheet extent and volume,
identification of the main ice-dispersal centres througout the last glacial
cycle, and understanding of final deglaciation patterns. However, there remains
considerable room for model improvements beyond the ones achieved in this
thesis. The foremost sources of error that remain unassessed in this thesis
include potential buttressing effects of the Laurentide ice sheet, changing
precipitation patterns with evolving topography, and regional variability of
the geothermal heat flux.

@@

Certainly the most important assumption made in this thesis is that of a
Cordilleran ice sheet evolution entirely disconnected from its much larger
neighbour, the \emph{Laurentide ice sheet}. Geomorphological evidence from the
suture zone indicates that, during the LGM, both Cordilleran and Laurentide ice
flow were deflected \citep{Kleman.etal.2010, Margold.etal.2013,
Margold.etal.2013a}, although it remains unclear for how long this suture
condition has lasted \citep{Jackson.etal.1997, Bednarski.Smith.2007,
Gowan.2013}.

Nevertheless, during most of the last glacial cycle, the two ice sheets have
been disconnected \citep{Kleman.etal.2010}. Thus, this assumption only affects
numerical reconstructions of the Cordilleran ice sheet during the LGM and
subsequent deglacation. Modelling results by \citet{Gregoire.etal.2012}
indicate that the Laurentide ice sheet buttress resulted in a thicker
Cordilleran ice sheet. However, modelling alone will neither elucidate the
nature nor the duration of the interaction between of the Cordilleran and
Laurentide ice sheets. For this, additional field constraints will be required.

@@

Another important source of error is the treatment of \emph{precipitation}, to
which
the numerical ice sheet model is highly sensitive (\CCLI). In all sensitivity
studies, we base precipitation patterns on current fields derived from
atmospheric
reanalyses. Although the distribution of precipitation has undoubtedly
been affected by past changes in global climate and the intermittent presence
of ice sheets and ice caps in the Northern Hemisphere, such changes are
difficult to assess. In the North American Cordillera, much of the continental
interior was presumed dryer than today due to the presence of ice caps on the
Pacific Coast ranges, reinforcing the current topographic barrier. However,
some regions were probably wetter as a side effect of modified atmospheric
circulation patterns.

These effects are typically analysed using atmospheric
circulation models. However, I have shown that different atmospheric reanalyses
--- circulation models using data assimilation techniques --- fail to agree
regarding the present-day pattern of precipitation over the modelling domain.
These discrepancies, which arise as a consequence of the highly mountainous
terrain, have dramatic effects on the ice sheet model response. Considering the
lack of observational data for modelling past climates, the increased
computational demand, and a decreased familiarity to processes at play, we must
expect that, for ice sheet simulations forced by different palaeo-climate
records, the uncertainty associated to the choice of atmospheric forcing
will be even higher. Alternative methods of coupling between ice sheet
evolution and precipitation changes need to be developed.

@@

I expect that inclusion of geothermal heat patterns will allow modelled basal
flow directions and distances to be compared with glacial lineations, which is
one of the main sources of geomorphological evidence used to reconstruct former
ice sheets \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}. For
instance, in our simulations, large portions of the Cordilleran ice sheet
remain cold-based throughout the deglaciation (\CCYC, Fig.~15). As cold-based
glaciers typically do not impact their beds significantly, this could indicate
that some of the lineation patterns we observe within these zones of consistent
cold-based deglaciation could be older features, formed during an earlier phase
of advance or retreat of the ice sheet. However, without a better control on
the geothermal heat flux, such analyses are perhaps too premature.

@@

In this thesis, I present numerical reconstructions of the former Cordilleran
ice sheet to a level of detail that has, to my knowledge, not been attained
before. Although mathematically uncertain, the coupled SIA-SSA approximation,
combined with massively parallel computing, allowed for ice sheet modelling at
large spatial scales, high spatial resolution, and long time scales
characteristic of ice sheet growth and decay.

From the outcome of sensitivity studies and the comparison of model output to
field constraints, I conclude the following:

@@

\item Although spatial patterns of glaciation in the North American Cordillera
are sometimes poorly understood, they can be largely explained by the
present-day distributions of temperature and precipitation, provided an
accurate treatment of seasonality, orographic precipitation, and daily
temperature variability in the model, and a simple temperature-index
accumulation and melt model (PDD).

\item The LGM Cordilleran ice sheet was a transient feature, out of balance
with contemporaneous atmospheric temperatures. Its maximum extent is the result
of a 50\,ka-long evolution. Peak glacial temperatures were lower than those
needed sustain a constant ice sheet geometry because the ice sheet growth was
finally inhibited by the deglacial transition to warmer temperatures .
Therefore, this maximum stage is not representative of average conditions, as
shown in the modelling by persistent ice caps throughout most of the last
glacial cycle.

\item Patterns of growth and decay of the Cordilleran ice sheet vary spatially
and temporally. Two-dimensional, conceptual models of growth and decay
\citep[e.g.][]{Fulton.1991, Margold.etal.2013a} do not encompass such
diversity. In order to decipher the complexity of geological evidence, the
Cordilleran ice sheet needs to be thought of as a three-dimensional body of
moving ice in perpetual transformation towards evolving climate conditions.
